Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-20 paper/paper.tex
saying POMDPs are cool

\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_overview_DMN.pdf}
  \caption{Neurobiological overview of the DMN with its key constituent nodes and their associated functional roles relevant in our functional account.
  \label{fig:neurobiological_overview_DMN}
\end{figure}
%
\subsection{The posteromedial cortex: global monitoring and information integration}
The midline structures of the DMN,
including the posteromedial cortex (PMC) and
the medial prefrontal cortex (mPFC),
are probably responsible for the highest turn-over
of glucose energy consumption \citep{raichle2001pnas}.
These metabolic characteristics go hand-in-hand with
functional neuroimaging analyses \citep{andrews2010} in
suggesting that the PMC and mPFC
potentially represent the functional backbone of the DMN.

@@

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Partially Observable Markov Decision Processes (POMDPs)}.
A POMDP formalizes a sequential decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL},
which we propose as a useful formal framework for DMN function.

Model-free RL is plausibly realizable in the human brain. Indeed,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
In our present model-free RL framework,
a rat might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

@@

% \subsubsection{The Bellman equation:  characterizing optimal behavior}
% Q-learning (Watkins \& Dayan, 1992) generates a
% \textit{greedy policy}
\subsubsection{Optimal behavior and the Bellman equation} Optimal behavior corresponds to a strategy $\pi^*$ for choosing actions such that, for every state, the chosen action guarantees the best possible reward on average. Formally,
% \textbf{XXX please add 1-2 sentences why this is interesting in the current
% context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) := \argmax_{a \in \mathcal A}{Q}^*(s, a),
  \label{eq:qlearning}
\end{equation}
where
\begin{equation}
  Q^*(s, a) := \max_{\pi}Q^\pi(s,a),
\end{equation}
the optimal Q-value over all policies. The goal of learning is to approach the policy $\pi^*$ as ``close as possible'', that is to \text{solve} the POMDP.
Of course \eqref{eq:qlearning} is just a definition and
represents by no means, a candidate schema for solving POMDPs with even moderately-sized action and state spaces: it's intractable!
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without querying the entire space of policies, viz
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the ``Bellman transform'' $\text{Bel}(Q)$ of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is defined by setting
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \gamma\mathbb E_{s' \sim p(s'|s,a)} [\max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a ``self-consistency'' equation which provides
a recursive decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
% The optimal Q-value operator $Q^*$
% is a fixed point for this equation.
As a consequence, the complicated dynamic programming optimization problem \eqref{eq:qlearning}
is broken into simpler sub-problems at different time points.
Exploitation of hierarchical structure in actions consideration
has previously been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

@@

\paragraph{Does hippocampal replay equal inverse reinforcement-learning?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  a POMDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For instance, given traces of motor actions from an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former?
  How? Can they reproduce this optimal behavior?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\cite{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

IRL is is suited for problems in which it's hard to define what the reward function could be
(e.g.,  car-driving, drone maneuvering, etc.) ...

>>>>

\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_overview_DMN.pdf}
  \caption{Neurobiological overview of the DMN with its key constituent nodes and their associated functional roles relevant in our functional account.}
  \label{fig:neurobiological_overview_DMN}
\end{figure}
%
\subsection{The posteromedial cortex: global monitoring and information integration}
The midline structures of the DMN,
including the posteromedial cortex (PMC) and
the medial prefrontal cortex (mPFC),
are probably responsible for the highest turn-over
of glucose energy consumption \citep{raichle2001pnas}.
These metabolic characteristics go hand-in-hand with
functional neuroimaging analyses \citep{andrews2010} in
suggesting that the PMC and mPFC
potentially represent the functional backbone of the DMN.

@@

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Partially Observable Markov Decision Processes (POMDPs)}.
A POMDP formalizes a sequential decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL},
which we propose as a useful formal framework for DMN function.
%

Model-free RL has had great success in real-world problems including robotics\cite{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering},
etc., just to name a few. Such a learning scheme is plausibly realizable in the human brain. Indeed,
a core property of human intelligence has been proposed underly to improve
on an expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of POMDPs. Also,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

In our adopted model-free RL framework,
a rat might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

@@

% \subsubsection{The Bellman equation:  characterizing optimal behavior}
% Q-learning (Watkins \& Dayan, 1992) generates a
% \textit{greedy policy}
\subsubsection{Optimal behavior and the Bellman equation} Optimal behavior corresponds to a strategy $\pi^*$ for choosing actions such that, for every state, the chosen action guarantees the best possible reward on average. Formally,
% \textbf{XXX please add 1-2 sentences why this is interesting in the current
% context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) := \argmax_{a \in \mathcal A}{Q}^*(s, a),\text{ where }Q^*(s, a) := \max_{\pi}Q^\pi(s,a).
  \label{eq:qlearning}
\end{equation}
The goal of learning is to approach the policy $\pi^*$ as ``close as possible'', that is to \text{solve} the POMDP.
Of course \eqref{eq:qlearning} is just a definition and
represents by no means, a candidate schema for solving POMDPs with even moderately-sized action and state spaces: it's intractable!
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without querying the entire space of policies, viz
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the ``Bellman transform'' $\text{Bel}(Q)$ of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is defined by setting
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \gamma\mathbb E_{s' \sim p(s'|s,a)} [\max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a ``self-consistency'' equation which provides
a recursive decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
% The optimal Q-value operator $Q^*$
% is a fixed point for this equation.
As a consequence, the complicated dynamic programming optimization problem \eqref{eq:qlearning}
is broken into simpler sub-problems at different time points.
Exploitation of hierarchical structure in actions consideration
has previously been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

@@

% \paragraph{Does hippocampal replay equal inverse reinforcement-learning?}
% %\begin{mdframed}
%   Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
%   a POMDP (called a \textit{teacher's demonstration}), can we figure out what is the
%   (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
%   \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
%   (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
%   For instance, given traces of motor actions from an adult to grab a cup on a table,
%   can an observing child figure out what ``langragian'' functional is being minimized by the former?
%   How? Can they reproduce this optimal behavior?
% %\end{mdframed}
% Such questions are of course pertinent to our decision-making theory for the DMN.
% In the general artificial case, the problem has been extensively studied and partially solved by
% \cite{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
% ``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

% IRL is is suited for problems in which it's hard to define what the reward function could be
% (e.g.,  car-driving, drone maneuvering, etc.) ...
